How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2024-03-15 2.03-NoHyp.Rmd
Add files via upload
# 2024 updates

So far, we have concentrated on thesis chapters or manuscripts that have [hypotheses](#hypothesis) to test. I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in context. But what, I hear you cry, if you don’t have a hypothesis to test?

There are lots of reasons why manuscripts don’t have hypotheses, and the chances are that in a PhD thesis (for example) you will have at least one chapter where there isn’t a hypothesis tested. So, given all the emphasis above, what can be done?

@@

## Does it matter that you don’t have a hypothesis?
Having a hypothesis makes your work stronger than if you only have a question or prediction. In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable. Of course, Popper would argue that if you don’t have a null hypothesis, then your methodology isn’t scientific.

To me, it is important for you to know exactly what your study is before you start writing about it, or even before you do the experiments. It is essential for you to frame the context of your study. The central reason why you are writing the study will lie at the heart of the manuscript. In the formulaic approach, it will be verbalised at the end of the introduction. All parts of the methodology will explain how you approached this central aim. So if you can’t pin down what this is, then make sure you discuss it with your advisor.

## Avoid HARKing 
Hypothesising After Results are Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting]. HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor. Increasing chances of confirmation bias, or a [Type I error](#TypeI), is not desirable as you are more likely to accept the alternative hypothesis when it is not correct. Alternatively, you should seriously consider [preregistering your aims](#transparency) to remain transparent.

>>>>

So far, we have concentrated on thesis chapters or manuscripts that have [hypotheses](#hypothesis) to test. I have emphasised that there is a need to focus all of the sections of your writing to help frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in the context of other work. But what if you don’t have a hypothesis to test?

There are lots of reasons why manuscripts don’t have hypotheses, and the chances are that in a PhD thesis (for example) you will have at least one chapter where there isn’t a hypothesis tested. It's also possible that you won't have any hypotheses tested, and this is the approach in some disciplines. So, given all the emphasis above, what can be done?

## Is having a hypothesis the gold standard?

The [hypo-deductive method](#whathypothesis) is certainly a powerful one to use in science but, no matter what you have read or been taught elsewhere, it is not the only scientific approach. Much of what we do in science comes from observational work, and these observations are usually the basis for many of the hypotheses that we pose. But some disciplines are entirely based on observations, and this does not make them obsolete or unscientific. Such disciplines simply don't lend themselves to experimentation, and without an experiment it is hard to control and hypothesise. But we should not confine ourselves to an entirely experimental approach. Here are some reasons why using non-experimental approaches will benefit our science:

1. Observational science (something that I refer to elsewhere as [Natural History Observations](https://howtopublishscience.org/whattowrite.html#natural-history-observations)) can teach us a lot more about systems and inspire greater creativity in our work than plodding from experiment to experiment. 
1. We can infer patterns from process. As humans, we are evolutionarily programmed to see patterns (even when they aren't there) and recognise patterns in the world around us [e.g., @linhares2013what]. It would be crazy to put this ability aside and ignore it when conducting science. Clearly, we are also able to make mistakes when we infer patterns from processes, but this does not mean that it should not be done. But we need to engage in a much more cautious manor when using this approach, always being aware of our bias towards certain explanations so that we encompass others. I would also argue here that this is instrumental in forming hypotheses to test experimentally or in models. 
1. Models of systems are powerful tools to test questions. Models are always simplified versions of complex biological systems, but their power as tools in our scientific approach should no longer be questioned.

Indeed, there are many shortcomings to using a hypo-deductive approach, as we have seen in the [previous chapter](#TypeI). By allowing ourselves a liberal approach to science methodology we are far more likely to make new connections and infer novel patterns about biological processes. We are more able to be inclusive of different disciplines. It is likely to be advantageous to be open to these diversity of approaches in others' research, just as you should in your own approach.

## Why am I always banging on about hypotheses?
Throughout this book (and in [other books](https://howtopublishscience.org/)) I constantly refer to hypotheses and how important they are when writing your thesis chapters. Why do I do this when I am clearly so positive about using other methods? In many respects, this comes from my attempt to simplify. In most cases, you could replace the word "hypothesis" with "central problematic" and get the same effect. You could also replace some instances with "aim" or "objective" (but not all). There are some points that are unique to hypotheses, and that's why I made my choice about referring to these as they are, generally speaking, the most complex of these alternative approaches. I reason that if you can understand how to approach writing to respond to hypotheses, then you can turn this same approach to your own needs. Add to this that most people, scientists included, misunderstand hypotheses, so there is a need to discuss them and their uses more.

If you are a biological scientist that does not use hypotheses, then this does not make you inferior, and I hope that you will still find this book useful. If this is you, then let me apologise here and suggest that you mentally shift whenever you read "hypothesis" to the alternative approach that you prefer.

@@

## Does it matter that you don’t have a hypothesis?
Proponents of the hypothetico-deductive method will argue that having a hypothesis makes your work stronger than if you only have a question or prediction. In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable. Of course, Popper would argue that if you don’t have a null hypothesis, then your methodology isn’t scientific.

## Avoid HARKing 
There is one very important aspect of your approach whether or not you have a hypothesis, and that is that you know exactly what your study aims are before you start. It is essential for you to frame the context of your study. The central reason why you are conducting the study will lie at the heart of the manuscript, once you've started writing. In the formulaic approach (see [Part III](#formula)), it will be verbalised at the end of the [introduction](#introduction). All parts of the [methodology](#methods) will explain how you approached this central aim. So if you feel that you still don't know what the objective of a certain chapter is, then make sure you discuss it with your advisor so that you know before you start.

Hypothesising After Results are Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting]. HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor. Increasing chances of confirmation bias, or a [Type I error](#TypeI), is not desirable as you are more likely to accept the alternative hypothesis when it is not correct. Alternatively, you should seriously consider [preregistering your aims](#transparency) to remain transparent.

### 2024-03-15 2.05a-AI.Rmd
Add files via upload
# 2024 updates

In this chapter I set out some ideas about how you might use AI to help you write, as well as some things that you should consider if you do go that route.

## What is a Large Language Model (LLM)?
I decided to ask ChatGPT and Bard to contribute to this chapter by describing themselves. My input text: "I am writing a chapter for a book on academic writing and I want you to describe ChatGPT in your own words". Here's what they said:

@@

@@

#### The results of my experiment are in - both bots failed {#AIfail}

@@

There will likely be better AI chat bots in the future, and ChatGPT itself is constantly getting feedback from users that should improve its own output.

@@

@@

>>>>

In this chapter I set out some ideas about how you might use AI to help you write, as well as some things that you should consider if you do go that route. Bear in mind that this field is moving extremely quickly and it will not be possible to keep up with every change happening in AI all of the time. Instead, I am to discuss more generically about how to use these tools.

## What is a Large Language Model (LLM)?
Back in February 2023, I decided to ask ChatGPT and Bard (as it was then) to contribute to this chapter by describing themselves. My input text: "I am writing a chapter for a book on academic writing and I want you to describe ChatGPT/Bard in your own words". Here's what they said:

@@

### Failure to learn
Lastly, a great reason not to use AI to help you write you thesis chapters is that you will learn next to nothing this way. Even if you aren't going to continue in science, writing a PhD is a significant learning experience, and you will benefit in being able to write reports and other types of documents since you will have had the experience of going from having nothing, planning what to write and then executing this. Don't underestimate the importance of this ability. The reality is that once you have accomplished writing your thesis you will have an important and marketable skill.

Using AI to generate text, even if it gets to the point in the future when it generates something useful, will impact your ability to generate text yourself. I can imagine a future where far fewer people are learning how to write because there is so much AI generated text. In this future, having the ability to write yourself will give you an advantage that should be valued.

@@

#### The results of my experiment are that both bots failed {#AIfail}

@@

In late 2023, an article in Nature [@conroy2023scientific] highlighted research by Guillaume Cabanac that had found the phrase "regnerate response" in more than 30 published papers. It is suggested that the authors have been so sloppy with their use of AI that they have also included the text on the button that occurs with the output. The examples spotted include the most crude errors and likely represent a very small proportion of AI generated text now published. The real question remains why so few people are critically reading manuscripts before they are published. Let us not forget that this is not the first time computer generated text has been published by "reputable publishers". An article from 2014 talks about how use of the crude text generation software *SCIgen* had been outed in dozens of published papers [@sample2014how].

Now (in March 2024) that I've had more than a year to evaluate AI tools that write text, I am still of the opinion that AI is not a solution, but can be used as a crude tool. In essays that students have submitted for assessment, **it has been easy to spot those who have used AI** to generate their text. How? Because AI gets the facts wrong, but still makes it sound good. These mistakes are really unlikely for a human to make. For example, AI will use real species names to summarise the results from an experiment, but uses the wrong name. The AI interpretation of what the experiment is about and what the findings are is often faulty, and can sometimes be wildly inaccurate. I suspect that students who were more diligent in what they handed in, were unlikely to have gained much advantage to using AI. Maybe it helped them get started, or maybe it helped them find some useful phrases, but there is still little help from generating text and using it without considerable reworking.

There will likely be better AI chat bots in the future, and ChatGPT itself is constantly getting feedback from users that should improve its own output. Hence, we can expect that current problems will change.

@@

### Universities are asking you to declare that text in your thesis is not AI generated
It would be worth checking with your institution to find out what their latest guidelines are. I have already received PhD theses to examine where there are declarations made that no text is AI generated. It is very likely that most institutions will make policies on this soon, if they have not done so already.

@@

### Publishers now make it a requirement that you declare no text was AI generated
Many publishers are now making it a requirement that you declare on submission of a manuscript that it was not AI generated.

### 2024-03-15 2.19a-concept.Rmd
Add files via upload
# 2024 updates

- Do you have hypotheses?

@@

## Your thesis timeline {#Gendt}

@@

### Producing a Gendt diagram
The timeline of your thesis will be an important guide throughout the 3 years that you are conducting your PhD. This means that it is worth getting this onto the record as soon as you are able. One of the standard ways of doing this is to produce a Gendt diagram. This takes each of the chapters along the entire timeline of your PhD and shows when you will work on each one. It may be that you want to break down each of these chapters into sub areas such as literature review, data collection, data analysis, write-up.

Below you will find a very simple version of a Gendt diagram (Figure \@ref(fig:Gendt)) that I put together for a fictional PhD. Note how the colours used for each chapter are the same as the thesis concept diagram (Figure \@ref(fig:concept)). It's always useful to follow such colour conventions once you've started them. Maintaining consistency throughout your thesis will help you when you explain it to others, such as during your proposal or whenever you present to your thesis committee.

(ref:Gendt) **A Gendt diagram used to help thesis plan how you will use your time during your PhD**. In this (fictional) example, I show how how the different chapters (Q1, T1, Q3, Q4, Q5) will take up different time during the three years of a PhD. Each year (1 to 3) is divided into four quarters (Q1 to Q4). This example is set out to show the need for seasonal sampling (in the second quarter of each year). Stars represent goals that you have set.

```{r Gendt, echo=FALSE, out.width = '95%',  fig.cap="(ref:Gendt)"}
knitr::include_graphics('figures/Gendt.png')
```

As you're doing your thesis in the biological sciences, you're much more likely to have some kind of seasonal quotient that constrains the work that you're doing. So in Figure \@ref(fig:Gendt) I've used the second quarter of each year as the key field period that is needed for all of the chapters.

@@

It's really useful to set down goals that you want to achieve on the Gendt plan, and to share these with your advisor. Milestones in your PhD are really useful when planning, and psychologically enhance your ability to finish different components.

@@

If you plan to attend workshops or conferences during your PhD (or even plan to organise these yourself), you should plan such significant events onto your timeline. For example, If you are presenting the results of your first chapter at a conference, you need to make sure that you have the necessary results well in advance of needing to write the abstract. In Figure \@ref(fig:Gendt), the first conference is in the third quarter of the second year and you should have already achieved your main goal for this chapter well in advance to write the abstract. You might also be able to add in the results of the second year's seasonal field work to present at the conference. Similarly, for the second conference, you should have all the results from the third chapter well in advance to write the abstract.

@@

In Figure \@ref(fig:Gendt) I have shown how a contingency might be mapped onto Chapters Q1, Q3 and Q4 with the potential addition of another field season in the second quarter of the third year in case one of the previous field seasons fails.

>>>>

- Do you have hypotheses, or do you have another approach?

@@

## Your thesis timeline {#Gantt}

@@

### Producing a Gantt chart
The timeline of your thesis will be an important guide throughout the 3+ years that you are conducting your PhD. This means that it is worth getting this onto the record as soon as you are able. One of the standard ways of doing this is to produce a Gantt chart. This takes each of the chapters along the entire timeline of your PhD and shows when you will work on each one. It may be that you want to break down each of these chapters into sub areas such as literature review, data collection, data analysis, write-up.

Remember that [planning  your time](#Plan) is empowering and something that you should be considering not only during your PhD (short-term), but also mid-term and long-term (see [Part I](#Plan)). To achieve your goals, you need to (a) know what they are, and (b) actively pursue them.

Below you will find a very simple version of a Gantt chart (Figure \@ref(fig:Gantt)) that I put together for a fictional PhD. Note how the colours used for each chapter are the same as the thesis concept diagram (Figure \@ref(fig:concept)). It's always useful to follow such colour conventions once you've started them. Maintaining consistency throughout your thesis will help you when you explain it to others, such as during your proposal or whenever you present to your thesis committee.

(ref:Gantt) **A Gantt chart used to help thesis plan how you will use your time during your PhD**. In this (fictional) example, I show how how the different chapters (Q1, T1, Q3, Q4, Q5) will take up different time during the three years of a PhD. Each year (1 to 3) is divided into four quarters (Q1 to Q4). This example is set out to show the need for seasonal sampling (in the second quarter of each year). Stars represent goals that you have set.

```{r Gantt, echo=FALSE, out.width = '95%',  fig.cap="(ref:Gantt)"}
knitr::include_graphics('figures/Gantt.png')
```

As you're doing your thesis in the biological sciences, you're much more likely to have some kind of seasonal quotient that constrains the work that you're doing. So in Figure \@ref(fig:Gantt) I've used the second quarter of each year as the key field period that is needed for all of the chapters.

@@

It's really useful to set down goals that you want to achieve on the Gantt chart plan, and to share these with your advisor. Milestones in your PhD are really useful when planning, and psychologically enhance your ability to finish different components.

@@

If you plan to attend workshops or conferences during your PhD (or even plan to organise these yourself), you should plan such significant events onto your timeline. For example, If you are presenting the results of your first chapter at a conference, you need to make sure that you have the necessary results well in advance of needing to write the abstract. In Figure \@ref(fig:Gantt), the first conference is in the third quarter of the second year and you should have already achieved your main goal for this chapter well in advance to write the abstract. You might also be able to add in the results of the second year's seasonal field work to present at the conference. Similarly, for the second conference, you should have all the results from the third chapter well in advance to write the abstract.

@@

In Figure \@ref(fig:Gantt) I have shown how a contingency might be mapped onto Chapters Q1, Q3 and Q4 with the potential addition of another field season in the second quarter of the third year in case one of the previous field seasons fails.

### 2024-03-15 3.03-Introduction.Rmd
Add files via upload
# 2024 updates

@@

@@

Remember that while this section on writing the introduction is short, much of [Part II](#part2) concerns writing style that is applicable to your introduction. In particular, sections on composing your [hypothesis](#hypothesis), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.

>>>>

### The hypothesis at the heart of the formula

Having made the point that the formula is funnelling the readers' interest, the question arises: to what? What are we working towards? Why are we so interested in having the reader focus on a particular thing, and more importantly: what is that thing?

> The Hypothesis

Your hypothesis is the reason that you are conducting your study. You are trying to respond to it, trying to accept or reject it. Your hypothesis is why you are doing the study and you need your reader to get the context for this, and to understand exactly why it was that you posed this specific hypothesis.

The hypothesis is then at the heart of everything that you write in your chapter (Figure \@ref(fig:formula-concept)):

 - Your introduction explains why you posed the hypothesis
 - Your methods & materials describe how you went about testing your hypothesis
 - Your results provide the outcome of the tests of your hypothesis
 - Your discussion responds to your hypothesis and provides the context

(ref:formula-concept) **Your hypothesis lies at the heart of your PhD chapter.** Your hypothesis (red star) is at the heart of everything that you are writing about. The entire reason for providing the chapter is to test your idea.

```{r formula-concept, echo=FALSE, out.width = '95%', fig.cap="(ref:formula-concept)"}
 knitr::include_graphics("figures/formula-concept.gif")
```

@@

I suggest that you keep your outline with you as you write and develop the introduction. It is important that you remember the objective of each paragraph. Sometimes, you can put it away, then when you bring it out again and start working on it you find that it changes into something else. At this point, you need to make a judgement on whether it has changed into something better. You can only really make this judgement if you remember what your original objective was. When you think that it has become a better paragraph, you need to check that it still meets your original objectives. If it doesn't, then you'll need to decide whether or not that objective was met elsewhere.

@@

### Should you include the results when introducing your study?

> How do you keep an idiot in suspense? I'll tell you later...

Increasingly, more scientists are including a concise statement during the last paragraph that responds to the hypothesis and states what the major outcome of the study is. The arguments for and against this have been made elsewhere [@heard2024scientific], and I will attempt to summarise them here:

**Against:**

 - This is not the convention in science 
 - Results do not appear in the introduction
 - You are giving away the result

**For:**

 - Science writing is not like writing a murder mystery novel
 - If it helps readers then we should do it
 - Convention is not a good reason for adherence

In this guide, I have emphasised that the [formulaic approach](#formula) to writing advocated here is to help you get started as a writer. One of your goals should be to experiment with your writing and if you think that moving away from the formula is going to improve clarity and readability, then I think you should go for it. Next, I think that you should elicit feedback from peers and your advisor. The formula I have presented is conventional, and that's deliberate. I am not going to tell you that it is the best way of writing your particular thesis chapter, and I would go so far as to say that shaking up convention is a very positive thing to do.

But if you are struggling with getting started, then I am going to continue to advise you to stick to the conventional way. As it's tried and tested you can be confident that this way is going to advance you towards your goal of writing your PhD.

## Summing up

The aim of your introduction is to provide the reader with all of the background information necessary to understand your study. Although the contents of this section on writing the introduction are somewhat brief, much of [Part II](#part2) concerns writing style that is applicable to your introduction. In particular, sections on composing your [hypothesis](#hypothesis) ([or not](#nohypothesis)), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.

### 2024-03-15 3.05-Results.Rmd
Add files via upload
# corrections after July 2021 Copy Edits
Add files via upload
# 2024 updates

This is perhaps the easiest section of your chapter or manuscript to write. However, there are still many mistakes that are made. Let’s start with the approach that you should take, and then go through some of the common mistakes.

## The correct approach - responding to your hypotheses
Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section (Figure \@ref(fig:results)). Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis. You should already have all the tests that you used in the methods section.

The results section often includes a lot of statistical testing, resulting in lots of numbers inside brackets within the text. This is all expected and fine, but consider consolidating multiple iterations into a table (if appropriate). And if your tests are already in a table, then they should not be repeated in your text. Instead, simply make your statement asserting the results and add “(see Table 1)”.

When describing the relationship between two variables, make sure that you state the direction of the effect. This means not simply stating that they have a significant relationship, but what that relationship is. This means stating that one increases with the other (a direct relationship) or one increases when the other decreases (an inverse relationship). Note that other types of variable relationships exist. Once you have described the result, either make reference to the Table or Figure where this can be seen, or describe the results of the test in brackets.

If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P. Only if P is less than α (alpha is usually 0.05) can you use the word ‘significant’ when describing the effect.

@@

@@

See sections on [tables](tables.html) and [figures](figures.html) below.

## Descriptive results
Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results. These normally convey the size of the dataset that entered into the rest of the analyses. If the size of the dataset varies by design in the methods, then you don’t need it here. But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results. In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses. Your text can then be broken into sections that correspond with your questions, unless you only have one, in which case further sections are not warranted.

>>>>

This is perhaps the easiest section of your chapter or manuscript to write. However, there are still many mistakes that are made. Here, I will start with the approach that you should take, and then go through some of the common mistakes.

## Description of your data
Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results. These normally convey the size of the dataset that entered into the rest of the analyses. If the size of the dataset varies by design in the methods, then you don’t need it here. But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results. It may be that your methods do not always produce predictable results. In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses.

Whenever you provide a summary of a variable (e.g. the mean), remember to include the variance; the range,  standard deviation, or standard error in brackets following the mean. If you collected a lot of variables, it may be more relevant to have them summarised in a table.

Once you have described the result, either make reference to the [Table](#tables), [Figure](#figures) or [Supplementary Information](#suppinfo) where this can be seen. If your tests are already in a [table](#tables), then they should not be repeated in your text. Instead, simply make a summary statement asserting the results and add “(see Table 1)”.

It is important to include mention of how much data was missing compared to your aim in the [Material and Methods](#methods) section. For example, an entire replicate may have been compromised, or a collection date might have been skipped for logistical reasons.

If you are [not testing an explicit hypothesis](#nohypothesis) your results will concentrate on the description of your data, and your analysis of the pertinent information that it contains. This may be done through tables, figures or models. If you modelled your data, you will need to show all models tested either in a table in the results, or in the [Supplementary Information](#suppinfo). You will also likely need to describe how you selected the best model and describe it. Although it is correct to describe your model here, do not try to interpret your findings. This is done in the [discussion](#discussion).

## Responding to your hypotheses
The easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and use them as subheadings for your results section (Figure \@ref(fig:results)). Once you have these subheadings, you simply assemble all of the results that address that hypothesis. You should already have all the tests that you used in the [Material and Methods](#methods) section. No additional tests should be listed in the results if they were not already described in the [Material and Methods](#methods) section.

@@

Remember that the most important aspect to your results is the size of any effect that you find and the confidence intervals, and **not** whether or not it is significant [@dushoff2019can]. This is not to say that significance should not be reported, this is important, and care should be taken to report the significance in a non-binary fashion [@dushoff2019can; @muff2021rewriting] - see [below](#binary).

The results section often includes a lot of statistical testing, resulting in lots of numbers inside brackets within the text. This is all expected and fine, but consider consolidating multiple iterations into a [table](#tables) (if appropriate).

When describing the relationship between two variables, make sure that you state the direction of the effect. This means not simply stating that they have a statistical relationship, but what that relationship is. For example, explaining that one increases with the other (a direct relationship), or one increases when the other decreases (an inverse relationship). Note that other types of variable relationships exist. When describing this relationship, you should also include the effect size (how much the relationship differs from no relationship), and the [power of your test](#increasing-the-statistical-power) (usually the number of replicates that allows you to avoid a [Type II error](#TypeI)).

### Moving away from binary significance {#binary}

Traditionally, we would say that only if P is less than α (alpha is usually 0.05) should you use the word ‘significant’ when describing the effect. However, the statistical community (including the biological community) is waking up to the need to replace the idea of the binary reporting of significance ("significant" or "not significant"), with wording that better reflects the reality of statistical testing " [see @muff2021rewriting]. The reason for this is that it represents a misunderstanding of statistical testing, and can be very misleading to those that read the results. Instead, we must appreciate that the position of our estimate of the P statistic lies somewhere along a continuum of potential results, and should be reported as such. The suggested change in language can be simply implemented as shown in Figure \@ref(fig:results).

(ref:replace-significance) **Help your reader to better interpret the meaning of your statistical testing by changing your language.** The figure shows some suggested language to use when writing about your results depending on the value of P that you find. These words should replace the binary outcome "significant" or "not significant" [see @muff2021rewriting].

```{r replace-significance, echo=FALSE, out.width = '45%', fig.cap="(ref:replace-significance)"}
knitr::include_graphics('figures/replace-significance.png')
```

While this approach succeeds in getting away from the significance dichotomy, there are still dissenting voices with the concept of P values representing continuous measures of evidence [@goodman1988evidence]. Another approach is to talk about the results of your significance tests in terms of the clarity that they provide about the relationship between the variables measured [@dushoff2019can]: whether they show a "statistically clear relationship" (aka significant), or "statistically unclear" (aka insigificant) [see @dushoff2019can for examples]. Here, I stress that if you have a lower value for P (e.g. equating to strong evidence) you haven't provided any additional evidence, you simply have more confidence in the non-compatibility of your data with your null hypothesis (H~0~). Your result is confined to the dataset that you are presenting. If you want additional evidence, then you need to run replicate experiments. Or, better still, find another mechanism in the biological sciences through which you can measure or test the same relationship.

There are better alternatives when it comes to statistical testing, but getting the inferential wording correct may still represent a challenge for non-statisticians.

If you have a statistical test with a P value, you do have to insert the results into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P. My suggestion is that you also consider exactly what you can say about your result, the effect size, confidence intervals and inferred relationship between the variables, as well as any statistical significance.

@@

See sections on [tables](#tables) and [figures](#figures) below.
